---
title: "Practicum 1"
output:
  pdf_document: default
  html_notebook: default
---

### Connect to Server

```{r}
# 1. Library
library(RMySQL)

# 2. Settings
db_user <- 'root' 
db_password <- '1290'
db_name <- 'birdstrike'
db_table <- 'courses'
db_host <- 'localhost'
db_port <- 3306

# 3. Read data from db
mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)
```

## Create Tables

```{sql connection=mydb}
DROP TABLE IF EXISTS Birds
```

```{sql connection=mydb}
create table Birds(
  Bird_ID INTEGER NOT NULL PRIMARY KEY AUTO_INCREMENT,
  Species VARCHAR(80) UNIQUE DEFAULT 'Unknown Bird',
  size ENUM('Small', 'Medium', 'Large', 'Not Available') DEFAULT 'Not Available'
);
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Aircraft
```


```{sql connection=mydb}
create table Aircraft(
  Aircraft_ID INT NOT NULL PRIMARY KEY AUTO_INCREMENT,
  Model VARCHAR(100) UNIQUE DEFAULT 'Not Available',
  Engines Numeric DEFAULT NULL
);
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Airport
```

```{sql connection=mydb}
create table Airport(
  Airport_ID INT NOT NULL PRIMARY KEY AUTO_INCREMENT,
  Airport_Name VARCHAR(100) UNIQUE DEFAULT 'Not Available',
  Origin VARCHAR(100) DEFAULT 'Not Available'
);
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Record
```


```{sql connection=mydb}
create table Record (
  Record_ID Numeric NOT NULL PRIMARY KEY,
  Airport_ID INT DEFAULT NULL,
  Aircraft_ID INT DEFAULT NULL,
  Airport_Name VARCHAR(100),
  Aircraft_Name VARCHAR(50),
  Origin VARCHAR(50),
  Species VARCHAR(50),
  Flight_Date DATE DEFAULT NULL,
  Airline VARCHAR(50) DEFAULT 'Not Available',
  Indicated_Damage ENUM('No damage', 'Caused damage', 'Not Available') DEFAULT 'Not Available',
  Flight_Impact ENUM('None', 'Aborted Take-off', 'Precautionary Landing', 'Engine Shut Down', 'Other', 'Not Available') DEFAULT 'Not Available',
  Phase ENUM('Take-off run', 'Climb', 'Approach', 'Landing Roll', 'Descent', 'Taxi', 'Not Available') DEFAULT 'Not Available',
  Remarks TEXT,
  Altitude NUMERIC CHECK(Altitude >= 0) DEFAULT NULL,
  Cost NUMERIC CHECK(Cost >= 0) DEFAULT NULL,
  Pilot_Warned ENUM('Y', 'N', 'Not Available') DEFAULT 'Not Available',
  Sky ENUM('No Cloud', 'Some Cloud', 'Overcast', 'Not Available') DEFAULT 'Not Available',
  Precipitation ENUM('Fog', 'Rain', 'Snow', 'Fog, Rain', 'Rain, Snow', 'Fog, Snow', 'Fog, Rain, Snow', 'Not Available', 'None') DEFAULT 'Not Available',
  Bird_ID INTEGER DEFAULT NULL,
  Number_Injured Integer CHECK(Number_Injured >= 0),
  FOREIGN KEY (Bird_ID) REFERENCES Birds(Bird_ID),
  FOREIGN KEY (Airport_ID) REFERENCES Airport(Airport_ID),
  FOREIGN KEY (Aircraft_ID) REFERENCES Aircraft(Aircraft_ID)
);
```



##Q3) READING DATA

```{r}
mydata = read.csv("C:\\Users\\Owner\\Downloads\\BirdStrikesData.csv")  # read csv file 
head(mydata)
```



##Remove non-essential data

```{r}
mydata$Wildlife..Number.struck = NULL
mydata$Wildlife..Number.Struck.Actual = NULL
mydata$Remains.of.wildlife.collected. = NULL
mydata$Remains.of.wildlife.sent.to.Smithsonian = NULL
mydata$Altitude.bin = NULL
mydata$Is.Aircraft.Large. = NULL
```



##Creating Airport Table

```{r}
library("dplyr")
head(mydata) %>% select(Airport..Name, Origin.State) 
```

```{r}
mydata1 <- mydata %>% select(Airport..Name, Origin.State) 
names(mydata1)[names(mydata1) == "Airport..Name"] <- "Airport_Name"
names(mydata1)[names(mydata1) == "Origin.State"] <- "Origin"
```

```{r}
dbWriteTable(mydb, "airport", mydata1, append = TRUE, row.names = FALSE)
```


##Creating Aircraft Table

```{r}
library("dplyr")
mydata2 <- mydata %>% select(Aircraft..Make.Model, Aircraft..Number.of.engines.) 
names(mydata2)[names(mydata2) == "Aircraft..Make.Model"] <- "Model"
names(mydata2)[names(mydata2) == "Aircraft..Number.of.engines."] <- "Engines"
head(mydata2)
```



#### if local data loading is disabled, use
#### mysql> show global variables like 'local_infile';
#### mysql> set global local_infile=true;
#### mysql> exit


```{r}
dbWriteTable(mydb, "aircraft", mydata2, append = TRUE, row.names = FALSE)
```


## Creating Bird Table

```{r}
library("dplyr")
mydata3 <- mydata %>% select(Wildlife..Species, Wildlife..Size) 
names(mydata3)[names(mydata3) == "Wildlife..Species"] <- "Species"
names(mydata3)[names(mydata3) == "Wildlife..Size"] <- "Size"
head(mydata3)
```

```{r}
dbWriteTable(mydb, "birds", mydata3, append = TRUE, row.names = FALSE)
```


## Remove Non Essential Columns

```{r}
head(mydata)
mydata$Wildlife..Size = NULL
mydata$Is.Aircraft.Large. = NULL
mydata$Aircraft..Number.of.engines. = NULL
mydata$Aircraft..Type = NULL
head(mydata)
```


## Fill in Record table and rename columns to be same as SQL table
```{r}
library("dplyr")
names(mydata)[names(mydata) == "ï..Record.ID"] <- "Record_ID"
names(mydata)[names(mydata) == "Airport..Name"] <- "Airport_Name"
names(mydata)[names(mydata) == "Aircraft..Make.Model"] <- "Aircraft_Name"
names(mydata)[names(mydata) == "Effect..Impact.to.flight"] <- "Flight_Impact"
names(mydata)[names(mydata) == "FlightDate"] <- "Flight_Date"
names(mydata)[names(mydata) == "Effect..Indicated.Damage"] <- "Indicated_Damage"
names(mydata)[names(mydata) == "Aircraft..Airline.Operator"] <- "Airline"
names(mydata)[names(mydata) == "Origin.State"] <- "Origin"
names(mydata)[names(mydata) == "When..Phase.of.flight"] <- "Phase"
names(mydata)[names(mydata) == "Conditions..Sky"] <- "Sky"
names(mydata)[names(mydata) == "Wildlife..Species"] <- "Species"
names(mydata)[names(mydata) == "Pilot.warned.of.birds.or.wildlife."] <- "Pilot_Warned"
names(mydata)[names(mydata) == "Cost..Total.."] <- "Cost"
names(mydata)[names(mydata) == "Feet.above.ground"] <- "Altitude"
names(mydata)[names(mydata) == "Number.of.people.injured"] <- "Number_Injured"
names(mydata)[names(mydata) == "Conditions..Precipitation"] <- "Precipitation"
```


## Reformat Date in R dataframe to be read into SQL
```{r}
library("tidyr")
mydata <- separate(mydata, Flight_Date, into = c("Date", "Time"), sep = " ")
mydata$Time = NULL
mydata$newdate <- strptime(as.character(mydata$Date), "%m/%d/%Y")
mydata$newdate <- format(mydata$newdate, "%Y-%m-%d")
names(mydata)[names(mydata) == "newdate"] <- "Flight_Date"
mydata$Date = NULL
mydata$Flight_Date <- as.Date(mydata$Flight_Date)
head(mydata)
```

```{r}
dbWriteTable(mydb, "record", mydata, append = TRUE, row.names = FALSE)
```


## Fill in record table with foreign keys for Airport
```{sql connection=mydb}
UPDATE record, airport 
SET
record.Airport_ID = airport.Airport_ID
WHERE record.airport_Name = airport.airport_name AND record.Origin = airport.Origin
```


## Fill in record table with foreign keys for Aircraft
```{sql connection=mydb}
UPDATE record, aircraft
SET
record.Aircraft_ID = aircraft.Aircraft_ID
WHERE record.aircraft_Name = aircraft.Model
```


## Fill in record table with foreign keys for Birds
```{sql connection=mydb}
UPDATE record, birds
SET
record.Bird_ID = birds.bird_ID
WHERE record.species = birds.species
```


## Drop columns for record after filling foreign keys
```{sql connection=mydb}
ALTER TABLE record
DROP COLUMN species;
```

```{sql connection=mydb}
ALTER TABLE record
DROP COLUMN Aircraft_Name;
```

```{sql connection=mydb}
ALTER TABLE record
DROP COLUMN Airport_Name;
```

```{sql connection=mydb}
ALTER TABLE record
DROP COLUMN Origin;
```


## Q4)
```{sql connection=mydb}
SELECT 
    record.airline AS Airline, COUNT(DISTINCT Record.record_id) as BirdStrikes
FROM
    Record
WHERE
    Record.phase = 'Take-off run'
        OR Record.Phase = 'Climb'
GROUP BY Record.airline
```


## Q5)
```{sql connection=mydb}
SELECT 
    airport.Airport_Name AS Airport_Name, COUNT(Record.record_id) AS BirdStrikes
FROM
    Record, Airport
    WHERE Record.Airport_ID = airport.Airport_ID
GROUP BY Airport.airport_Name
ORDER BY BirdStrikes DESC;
```


## Q6)
```{r}
query6 <- dbGetQuery(mydb, "SELECT EXTRACT(YEAR FROM Record.flight_date) as Year, 
                     COUNT(DISTINCT Record.record_id) AS BirdStrikes FROM Record GROUP BY 
                     EXTRACT(YEAR FROM Record.flight_date);")
head(query6);
```


## Q7) Create View For ascent/Take-off Run

``````{sql connection=mydb}
DROP VIEW IF EXISTS ViewAsc
```


```{sql connection=mydb}
CREATE View ViewAsc AS
SELECT 
    EXTRACT(YEAR FROM Record.flight_date) AS YEAR,
    Record.Phase,
    COUNT(DISTINCT Record.record_id) AS BirdStrikes
FROM
    Record
WHERE EXTRACT(YEAR FROM Record.flight_date) >= 2008
GROUP BY EXTRACT(YEAR FROM Record.flight_date) , Record.Phase
HAVING Record.phase = 'Take-off run'
    OR Record.phase = 'Climb'
```

``````{sql connection=mydb}
DROP VIEW IF EXISTS ViewDesc
```


## Create view for Descent/Landing Roll
```{sql connection=mydb}
CREATE View ViewDesc AS
SELECT 
    EXTRACT(YEAR FROM Record.flight_date) AS YEAR,
    Record.Phase,
    COUNT(DISTINCT Record.record_id) AS BirdStrikes
FROM
    Record
WHERE EXTRACT(YEAR FROM Record.flight_date) >= 2008
GROUP BY EXTRACT(YEAR FROM Record.flight_date) , Record.Phase
HAVING Record.phase = 'Landing Roll'
    OR Record.phase = 'Descent'
```


## Import Ascent/Take-off data into data frame

```{r}
chart1 <- dbGetQuery(mydb, "Select Year, sum(BirdStrikes) as BirdStrikes from viewAsc Group By Year");
chart1["Phase"] = "Ascent + TakeOff"
chart1
```


## Import Descent/Landing data into data frame

```{r}
chart2 <- dbGetQuery(mydb, "Select Year, sum(BirdStrikes) as BirdStrikes from viewDesc Group By Year");
chart2["Phase"] = "Descent + Landing"
chart2
```


## Merge the two

```{r}
chart3 <- rbind(chart1, chart2)
chart3
```


## Create Graph

```{r}
library("ggplot2")
ggplot(data = chart3, mapping = aes(x = Year, y = BirdStrikes, fill = Phase)) + 
  geom_bar(stat = "identity", position = "dodge")
```


## Q8)

```{sql connection=mydb}
DROP Procedure IF EXISTS deleteRecord
```

```{sql connection=mydb}
CREATE DEFINER=`root`@`localhost` PROCEDURE `deleteRecord`(IN recordID INTEGER)
BEGIN
Delete from record
where record.record_id = recordID;
END
```


#### Might show up as 0 rows since it has already been deleted. Re-read data into SQL or enter new Record_ID

```{sql connection=mydb}
Select * from record where Record_ID = 205958
```

```{sql connection=mydb}
Call deleteRecord(205958);
```

```{sql connection=mydb}
Select * from record where Record_ID = 205958
```


## Disconnect from Database

Disconnect from the database (important as database servers have limited numbers of connection and each connection uses a resource).

```{r}
dbDisconnect(mydb)
```

